Unknown

### 2024-03-06 ms/ms.qmd
working on materials and methods

@@

@@

outline of tab:accessions LA number,Species name, lat, lon, elev, SPLASH light value

>>>>

percent_terminal = "70.0" # percent of LICOR measurements on terminal leaflet
percent_lateral = "30.0"
percent_change_leaflet = "10.0" # percent of time that anatomical traits were measured on different leaflet than LICOR measurements
```

@@

```{r, accessions, eval=FALSE,echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-accessions
#| tbl-cap: "Solanum accessions"

# outline of tab:accessions LA number,Species name, lat, lon, elev, SPLASH light value

# copied code from stomata-ilima to emulate
# read_rds("../processed-data/site.rds") |>
#   filter(site_code != "nnpl") |>
#   mutate(Island = case_when(
#     island == "oahu" ~ "Oʻahu",
#     island == "hawaii" ~ "Hawaiʻi"
#   )) |>
#   arrange(site_type, site) |>
#   select(Site = site, Island, Habitat = site_type, Latitude = latitude_degree,
#          Longitude = longitude_degree, `Elevation (mas)` = elevation_m) |>
#   knitr::kable(format = "latex", digits = c(rep(0, 3), 3, 3, 0),
#                booktabs = TRUE) |>
#   kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
#   kableExtra::column_spec(1, width = "4cm") |>
#   kableExtra::column_spec(2, width = "1.5cm") |>
#   kableExtra::column_spec(3, width = "1.5cm") |>
#   kableExtra::column_spec(4, width = "1.5cm") |>
#   kableExtra::column_spec(5, width = "1.5cm") |>
#   kableExtra::column_spec(6, width = "1.5cm")

@@

### Leaf trait measurements

We selected a fully expanded, unshaded leaf at least six leaves above the cotyledons during early vegetative growth. This typically meant that plants had grown in light treatments for $\approx$ 4 weeks, ensuring they had time to sense and respond developmentally to the light intensity of the treatment rather than the seedling conditions [@schoch_dependence_1980]. Shade plants grew slower than sun plants, hence leaves at the same developmental stage were measured on chronologically older plants in the shade treatment. In some sun plants, we had to use leaves higher on the stem because short internodes made lower leaves inaccessible with the gas exchange equipment. We measured terminal leaflets in `r percent_terminal`\% of cases, but used the lateral leaflet closest to the terminal leaflet when it was damaged or difficult to clamp into the gas exchange chamber. When a leaflet was damanged during gas exchange measurements, we collected anatomical data from the nearest leaflet on the same leaf (`r percent_change_leaflet`)\%. We measured chlorophyll concentration index (CCI) using a chrolophyll concentration meter (MC-100, Apogee Instruments, Logan, Utah, USA) on the lamina of focal leaflets before gas exchange measurements at the same time we measured leaf temperature.

### Amphistomy advantage

We estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024. $\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total $g_\text{sw}$:

$$\mathrm{AA} = \mathrm{log}(A_{\mathrm{amphi}} / A_{\mathrm{hypo}})$$

We measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values. We refer to this as an $A \textendash g_\text{sw}$ curve. We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).

We measured $A \textendash g_\text{sw}$ curves using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA). Light-acclimated plants were placed under LEDs dimmed to match their light treatment during gas exchange measurements. We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at ambient CO$_2$ (415 ppm) and $T_\mathrm{leaf} = \qty{25.0}{\degreeCelsius}$. The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission. To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm. We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.

We collected four $A \textendash g_\text{sw}$ curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve at high light-intensity ($\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and low light-intensity ($\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$). We always measured high light-intensity curves first because photosynthetic downregulation is faster than upregulation in these species. To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements. Unlike @triplett_amphistomy_2024, preliminary experiments with *Solanum* indicated a strong order effect in that $A$ declined in the second curve. Therefore, we made measurements over two days. On the first day, we measured high and low light-intensity curves for either amphi or pseudohypo leaves; on the second day, we measured high and low light-intensity curves on the other leaf type.

In all cases, we acclimated the focal leaf to high light ($\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and high relative humidity ($\mathrm{RH} = 70\%$) until $A$ and $g_\text{sw}$ reach their maximum. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation. Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then acclimated the leaf to low light ($\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and 70\% $\mathrm{RH}$ before inducing stomatal closure with low $\mathrm{RH}$.
